Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-03-09 paper/manuscript.Rmd
edits to discussion, removed most predator stuff, added a couple sentences on non-ice-free movements that need numbers

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond2016]. We present two
explanations for these swimming events, forage limitation [@Miller2002] and
predator avoidance [@Jeffery2007], and discuss why these swimming events
appear to be relatively commonplace in the population.

@@

**Predator avoidance**

Predator avoidance could also contribute to swimming behaviour of caribou
[@Miller2002]. During calving, female caribou are known to swim to small islands
in freshwater lakes and give birth, or with their calves after birth, to avoid
predation [@Bergerud1985]. The use of shoreline habitat in summer is considered
an effective anti-predator calving strategy by allowing for swimming to islands
or peninsulas as a predation avoidance flight response [@Bergerud1990]. The use
of water, and by association occasionally swimming, is therefore part of the
fine-scale interactions between caribou and their predators and the use of
islands appears to be an effective anti-predator strategy.

Historic predators of caribou such as wolves (*Canis lupus*) and black bears
(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
adult caribou is relatively rare [@Bastille-Rousseau2016], so the potential
effect of predators on caribou swimming behaviour is limited to non-consumptive
effects associated with encounters and subsequent flight response. Adult females
with calves at heel may be an exception because calves are highly susceptible to
coyotes [@Rayl2014], but the energetic costs for calves to swim in the ocean is
likely far greater than adults. As one might expect, none <!--confirm--> of the
86 swimming events we observed occurred in the weeks after calves were born,
suggesting swimming with in the ocean with a calf-at-heel is likely more risky
than other predator avoidance behaviours.

<!-- did the swimming events occur during the calving season? were these three breeders/non-breeders? did any of these females make these journeys with their calves?  -->

@@

In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occurring
once every 24 <!--confirm--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. <!-- QW: not sure how to finish this off -->
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

The forage limitation and predator avoidance hypotheses are generally thought to
apply uniformly across populations, but individual caribou could evaluate the
costs and benefits of swimming to another island differently. It is possible
that some individuals, in our case ~10%, considered swimming less costly or that
they perceived competition or predation more acutely than their conspecifics.
Intrinsic life-history, physiological, or behavioural traits that might
influence how animals vary in their assessments of swimming trade-offs remains
unknown. Future work could assess the role of age, stress, or an individuals
tendency to move potential predictors of swimming behaviour.

@@

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation to a series of proximate observations, fine-scale density-dependent
habitat selection is theoretically possible [@Webber2018]. Predator avoidance
could also contribute to caribou swimming behaviour, but given the presumed low
density of coyotes on Fogo Island and that predation primarily only affects
young calves, we suspect predator risk is a possible, but less likely
explanation for swimming.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit - how is this? -->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Taken
together, we suggest that swimming is likely more common than previously thought
for caribou living on oceanic islands. We posit forage limitation and the
associated density-dependent habitat selection is an ultimate explanation for
why caribou swimming in the ocean, with predator avoidance as a potential
proximate explanation. Although our inference is limited to observations, our
observations suggest that islands represent discrete foraging patches for
terrestrial animals that could influence fitness via increased foraging
opportunities and safety from predators. We also suggest that individuals vary
in their evaluation of costs and benefits of movement between these habitat
patches.

>>>>

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond2016]. We propose that one such
explanation, the forage limitation hypothesis, is likely driving these oceanic
swimming events [@Miller2002].

Predation pressure is also known to affect caribou swimming behaviour, both in
immediate situations to escape predators [@Bergerud1990] and also pre-emptively
to give birth or raise calves in safer areas like isolated islands
[@Bergerud1985]. The Fogo Island archipelago does not contain historic caribou
predators such as wolves (*Canis lupus*) or black bears (*Ursus americanus*),
leaving coyotes (*Canis latrans*) as the top predator. Coyotes may predate
calves [@Rayl2014] but do not pose a substantial risk to adult caribou
[@Bastille-Rousseau2016]; predation in general does not exert as strong a force
on Fogo Island caribou as in mainland populations. Calves that are most
vulnerable to predation likely face a much higher energetic cost of oceanic
swimming than do adults, and indeed none of the females observed swimming had a
calf with them at the time.<!-- female-centric rather than time-of-year
["swimming events occurred in the weeks after calves were born"] seems better,
if we can confirm this--> Given the frequency of these swimming events, and that
swimming in the ocean with a calf-at-heel is riskier than other predator
avoidance strategies, we suspect that predation risk is not a main driver of
oceanic swimming in Fogo Island caribou.

@@

@@

In total, 12 of 29 collared female caribou swam between islands in the Fogo
archipelago. Only 3 of these 12 had more than XX swimming events during the
study period, so approximately one third of our marked population swam
irregularly and infrequently, whereas for ~10% of individuals these swimming
events were relatively common, on average occurring once every 24 <!--confirm-->
days during ice-free seasons. We did not detect any swimming events for the
remaining 17 individuals; it remains unclear whether a subset of the population
does not engage in this behaviour at all, or whether these swimming events are
extremely rare for most caribou and we happened to detect some for those nine
individuals with very few observed swimming events. We are also unable to draw
conclusions about the prevalence of swimming in the ocean by male caribou: our
initial visual observation was of an adult male, but our GPS data only included
females.

<!-- "During the winter, when individuals could theoretically move between islands more easily by walking across ice, XX individuals were observed moving between islands. YY [(a) majority, (b) minority] of these were not detected swimming during the ice-free season and only moved between islands overwinter. This suggests that [(a) moving between islands is a relatively consistent behaviour within individuals, where some caribou refrain from both swimming and walking over ice] / [(b) caribou who avoided swimming between islands would still take advantage of the benefits provided by these patches, but only when travelling to them was less risky and costly] -->

<!-- QW: I'm not sure we need to say the below.. 
JGH - yeah, it's pretty extra I guess. I'm not really a fan of the above para either, it's not very convincing and doesn't add much
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

Regardless, oceanic swimming is relatively common among a small minority of our
marked population. Of the three collared individuals, two were collared on
Western Indian Island and the third was collared on Fogo Island. Inter-island
distances between Western and Eastern Indian Islands, and the smaller islands
around them, are shorter than those between Fogo Island and others (Figure 2).
Individuals that were born on or spend more time on these smaller islands may
swim between them more frequently than animals that spend most of their time on
Fogo Island. The forage limitation hypothesis is generally thought to apply
uniformly across populations, but individual caribou could evaluate the costs
and benefits of swimming to another island differently. It is possible that some
individuals, in our case ~10%, considered swimming to be less costly or
perceived competition or predation more acutely than their conspecifics.

@@

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation to a series of proximate observations, fine-scale density-dependent
habitat selection is theoretically possible [@Webber2018]. Given the presumed
low density of coyotes on Fogo Island and the minimal risk they pose to all but
young calves, who are least likely to engage in oceanic swimming, predation risk
does not appear to be a major driver of the behaviour in this system.
<!-- QW: changed it a bit - how is this? 
JGH: simplified these two para's a bit, took out most of the predator stuff. I'm a bit confused by the sentence above citing your 2018 paper, I'm not really sure what it's saying & how it relates to the rest of the section. there were comments to that effect as well, maybe expand the end a bit to make it clearer how that relates? -->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Taken
together, we suggest that swimming is likely more common than previously thought
for caribou living on oceanic islands, and that forage limitation and the
associated density-dependent habitat selection is an ultimate explanation for
this phenomenon. Although our inference is limited to observations, islands
appear to represent discrete foraging patches for terrestrial animals that can
influence fitness via increased foraging opportunities. We also posit that
individuals can vary in their evaluation of costs and benefits of movement
between these habitat patches.
